A comparison of different methods for calculating tangent-stiffness matrices in a massively parallel computational peridynamics code.

### 2013-09-03 jacobian_calc_CMAME.tex
implementation section, other additions

\subsubsection{What is a Tangent-stiffness Matrix}

@@

@@

\subsubsection{Justification: Goals, Assumptions, and Metrics} 
A goal of the study was to rank complex-step against forward-difference and central difference on
the basis of accuracy. AD was ommited from the comparison because it served as the standard of
accuracy for the other methods in the absence of appropriate analaytical forms for the Jacobian
associated with the nonlinear system solved in the study. The assumption that AD is accurate
enough to serve as a standard is supported by AD's implementation as a computerized chain rule as
explained in \ref{ADsubsection}. For a concrete comparison of each of the methods including AD to a
known analytical Jacobian, refer to the corresponding author's website, specifically the C++
source-code example \emph{BeamAndSpringSystem}.

@@

\subsubsection{Materials: Computers, Other Software}
Mesh generation for the test problems run in the study was done using the software package $Cubit$
developed at REFERENCE.

\subsubsection{Methods: Reduction of Data}

@@

\section{Conclusions and Further Work}
\subsection{Thoughts on Results}
\subsection{Thoughts on \emph{Perdigm}}
\subsection{Final Thoughts on Complex-Step}

>>>>

\subsubsection{What is a Tangent-stiffness Matrix}
\label{tsmat}

@@

Complex step was implemented in \emph{Perdigm} as follows. First a way of persistently allocating
memory needed to store imaginary values for nodal position and dilatation was determined. Persistent
memory allocation is a very basic way to improve performance in general for a computer program. The way
memory was allocated for similar vector and scalar variables used by the prior existing forward
difference method was copied and new 'field spec' variables were added into the 'elastic' material
model to store imaginary position coordinates and imaginary dilatation. Model evaluation computer
code methods that were used in the course of applying the forward difference method were copied,
renamed and re-written to follow the CTSE formula, which mainly involved changing the datatype of
intermediate variables which were method-local and dynamically casting process-local variables to
complex compatible datatypes. Additional modifications were for the purpose of collecting
experimental data and are not part of complex-step itself. The mathematical difference between the
original forward-difference and complex-step code was covered in \ref{tsmat}. The source code for
implementing complex-step in \emph{Perdigim} can be found on the corresponding author's website. Because
\emph{Peridigm} was written by computer scientists, many advanced techniques in C++, such as
multiple virtual inheritance, pointer arithmetic and distributed data structures were used in the
basic code and by necessity in the modified code. To see complex-step in a more narrow context, it
is advised that readers also take a look at the examples of complex-step provided on the
corresponding author's website.

@@

The methods were compared by running two sets of test problems where each method would solve a
problem concurrently. Each test problem simulated a $4 by .25 by .25$ meter block of metal
undergoing tension along the axis parallel to the long dimension of the block. The test problems
were set up in \emph{Perdigm} as qausistatic equillibrium problems, where the displacement boundary
conditions used to apply tension were applied gradually, in 'load steps', and an equilibrium
solution for each load step was achieved before applying the next load step. The material properties
of the model were set at $8.0E-9 metric tonne/ mm^3$ density, $1.515E4 MPa$ bulk modulus, and
$7.813E4 Mpa$ shear modulus. The peridynamic horizon used in a test problem was three times the node
spacing for the corrisponding model mesh. This number was chosen out of experience to insure that
every node could have at least one nonlocal connection along the diagonal of the initial node grid.

A series of nine of these test problems were run, with the mesh discretization used in the model
finer and finer with every test in the series. The aim of increasing the mesh complexity was to see
if possible differences between the methods in terms of accuracy and speed would be affected. This
series would comprise the single core runs. The specific parameters of these tests can be found in
\emph{Peridigm} input xml files included in the archive for this paper which can be found in the
corresponding author's website. These xml files allow users with appropriately modified versions of
\emph{Peridigm} to reproduce the results found in this paper.

Another series of four test problems were run, however this time the number of cores used to solve
the problem was varied from test to test rather than the mesh refinement. While mesh refinement was
not varied from test to test, it was set at a somewhat high $1E6$ perdidynamic nodes of
discretization. This allowed the simulation computer to be challenged sufficiently that the effect
of varying the number of cores used could be more clearly seen. The specific parameters of these
tests can be found in \emph{Perdigm} input xml files also included in the archive for this paper on
the corresponding author's website.

What was recorded during the test runs as experimental data is the subject the following
subsubsection. It is organized to eplain "what", "why" and "how" in that order.

\subsubsection{Justification: Goals, Assumptions, and Metrics} 
\label{JGAM}
A goal of the study was to rank complex-step against forward-difference and central difference on
the basis of accuracy. AD was ommited from the comparison because it served as the standard of
accuracy for the other methods in the absence of appropriate analaytical forms for the Jacobian
associated with the nonlinear system solved in the study. The assumption that AD is accurate
enough to serve as a standard is supported by AD's implementation as a computerized chain rule as
explained in \ref{ADsubsection}.

@@

\subsubsection{Materials: Computers, Other Software}

Mesh generation for the test problems run in the study was done using the software package
\emph{Cubit} developed at REFERENCE. For the test series in which peridynamic node density was
varied, journal scripts compatible with \emph{Cubit} were used. The rule used to select the
arangement of nodes within the borders designated by simulated material block's dimensions was to
manually choose a node spacing, fit the maximum whole number of nodes in a regular grid within the
block, and then check whether or not this number of nodes matched the number of nodes desired for a
particular test. The same process was used for the multicore series. An example journal script can
be found on the corresponding author's website. For future work, it may be beneficial to automate
and randomize this process for statistic reasons and for time efficiency.

The test problems were run the \emph{Stampede} HPC cluster computer housed at \emph{TACC} (Texas
Advanced Computing Center) based out of UT Austin. The single core test series was run using the
normal queue compute nodes, while the multicore test series was run using one to four "large memory"
nodes each having one terrabyte of local RAM. This large amount of memory was required for reasons
explained in \ref{JGAM}.

\subsubsection{Methods: Reduction of Data} The output produced by \emph{Perdigm} including the
additional accuracy and speed measurements it was modified to provide was redirected from the
console to text files by the resource manager, \emph{SLURM} (Simple Linux User Resource Manager).
Directories and filenames were chosen so that data would be easy to sort by individual test run.
Extraneous information was manually deleted from copies versions of the output text files so that
they would be regular enough to be easily read by scripts written in \emph{Python}. These scripts
averaged accuracy or speed data for a test corresponding to a particular peridynamic node density
over all iterations for that run, and then plotted this averaged data as if it were a function of
peridynamic node density as it varied from test to test in the series.  Similarly, for the multicore
tests, data was averaged and plotted in much the same way, but as a funciton of number of cores used
to solve the problem rather than as a function peridynamic node density which was held constant.
These data reduction and plotting scripts can be found on the corresponding autor's website.

Since samples were not repeated, because there was not resources for it, confidence intervals were
not calculated. However, if results of iterations within the same test problem were considered
samples of the same thing, then a confidence interval could be calculated. This assumption was made
for the purpose of plotting reporting average values, but a confidence interval was not calculated.
But, a cursory review of data showed no range overlap among the methods for accuracy or speed data
from any test run of either series.

@@

\section{Conclusions and Further Work}

\subsection{Thoughts on Results}

\subsection{Thoughts on \emph{Perdigm}}

After the bulk of the study had been conducted, it was determined that the implementations of
forward-difference, central-difference and by inherited code complex-step, in \emph{Peridigm} were
not optimal. Each of these implementations included extensive looping over noncontiguous addresses
in memory and unnecessary re-calculation of constant values. Unpublished preliminary results show
that when calculations for the quantity 'zeta' are re-used and only perturbed and previously
pertubed functions of independent variables are recomputed throughout the course of a Jacobian
evaluation, complex-step performs at the speed of forward-difference. Because of this it is
speculated that forward-difference and central-difference could be modified in nearly the same way to
perform more quickly than AD. An enhanced performance implementation of forward-difference will
eventually be made available on the corresponding author's website.

\subsection{Final Thoughts on Complex-Step}

Complex-step was shown to be highly accurate by the standards of this study, but as it is
implemented in \emph{Peridigm} may not be any better than AD, and it is almost certainly slower for
a naive implementations. When looking to the future of \emph{Peridigm}, Complex-step holds the
advantage of relying on byte-copiable datatypes, while AD requires a complicated serialization and
deserialization in order to function as byte-copiable. This is fine for copying in between MPI
processes, but deserialization on an accelerator or GPU is very difficult, as this would require
compiling a version of the AD library being used that were compatible with the accelerator or GPU
being used. What this means is that if \emph{Peridigm} were to be reformulated to take advantage of
accelerators like GPU's or similar, which is likely, given their performance and power efficiency
advantages over traditional host CPU's, complex-step would become more attractive.
